---
title: "EDS Raster workshop 2: analyzing with rasters"
author: "Casey O'Hara"
date: "11/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messages = FALSE, warnings = FALSE)

### load packages.  Since raster has a "select" function but we more often
### use the tidyverse::select, load raster first, and let tidyverse overwrite
### that function.

library(raster)
library(tidyverse)
library(sf)
library(fasterize)
library(here)
```
 
# Summary
 
The previous script focused on creating and developing rasters in preparation for analysis.  Here we'll do a little more , but focus on different ways you can use rasters to perform an analysis.

# Basic raster math

Since rasters are just a grid of numeric values, we can do the same operations on a raster as we can with a basic vector.  Here we'll walk through some basic but totally useless examples based on the rasters we already created.

```{r}
### operating on a vector
vec <- 1:10
vec * 2  ### multiply by a scalar
vec + 20 ### add a scalar

eez_rast <- raster(here('_spatial/se_afr_eez_6933.tif'))
plot(eez_rast)
eez_rast     ### note the values: 1 to 9
eez_rast + 5 ### note the values: 6 to 14
eez_rast * 2 ### note the values: 2 to 18
plot(eez_rast^3)
### note, plotting linear transformations will look exactly the same, since R
### chooses the palette to fill out the entire range.  The numbers on the 
### colorbar would change though!
```

We can also add, subtract, multiply, or divide by a raster with the same parameters.

```{r}
hh_rast <- raster(here('_spatial/hh_rast_6933.tif'))
plot(hh_rast)
plot(hh_rast + eez_rast)
```

You can also use indexing to change value in one raster based on the value in another raster.  Use the `raster::values()` function to basically convert the raster values into a vector that's less ambiguous to work with.  We already did this to fill in gaps in the silky shark range.
```{r}
x <- values(eez_rast)
unique(x) ### each number corresponds to an EEZ; Mozambique is 5
hh_rast2 <- hh_rast ### copy it over

### keep all shark probabilities in Moz, but set all outside the Moz EEZ to -1
values(hh_rast2)[values(eez_rast != 5) & !is.na(values(hh_rast))] <- -1
plot(hh_rast2)

### could also set non-Moz EEZ values to NA, basically applying a mask.
```


How can you use this basic raster math?

* basic combinations of rasters to quickly calculate ratios, sums, etc.
* rescale a raster by dividing by the max value `rast_rescale <- rast / maxValue(rast)`
* use a raster of presence/absence (as ones and zeros) to turn on or turn off values in a different raster (similar to a mask)
* "flatten" a raster to all ones and NAs by dividing the raster by itself (something divided by itself = 1, 0 divided by 0 = NaN)

```{r}
hcaf_rast <- raster(here('_spatial/hcaf_rast_6933.tif'))
hcaf_rescaled <- hcaf_rast / maxValue(hcaf_rast)
plot(hcaf_rescaled)
```


# `raster::calc()`

The `calc` function is handy for more complex calculations.  It is especially useful when you have a stack of raster layers and want to calculate across them all at once.  A raster `stack` object is just multiple layers stacked together.  A raster `brick` is similar.  At a basic level where we are now, the difference is pretty unimportant.

```{r}
### tell it a vector of files to include in the stack.
rast_files <- list.files(here('_spatial'), pattern = '_6933.tif', full.names = TRUE)
rast_stack <- stack(rast_files)
plot(rast_stack)

### or tell it which rasters already in memory to include in the stack.
rast_stack <- stack(eez_rast, hh_rast, hcaf_rescaled)

rast_mean <- calc(rast_stack, fun = prod, na.rm = TRUE)
plot(rast_mean)
```



- reclassify
- zonal

- values to create a dataframe


```
### distance raster example
dist_rast <- raster::distance(se_afr_eez_6933_rast)
plot(dist_rast)

### cell ID example
cell_id_rast <- se_afr_eez_6933_rast ### initialize it with a raster with the right resolution, CRS, and extent
values(cell_id_rast) <- 1:ncell(cell_id_rast) ### put in unique numbers for cell values
plot(cell_id_rast)

### raster to polygon example
hcaf_moz_poly <- raster::rasterToPolygons(hcaf_moz_rast, dissolve = FALSE)
se_afr_eez_poly <- raster::rasterToPolygons(se_afr_eez_rast, dissolve = TRUE)
plot(se_afr_eez_poly)
```

